Measurement Bias from Address Aliasing

### 2016-01-07 paper.tex
Update for iWAPT2016. Basic changes for compiling only.

% paper.tex
% ACM sigmetric format
% Author: L. K. Melhus and R. E. Jensen
\documentclass{acm_proc_article-sp}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{filecontents}
\usepackage{tikz}
\usepackage{listings}
\usepackage{mathtools}
\usepackage{underscore}
\usepackage{url}
\usepackage{lmodern} % Much clearer text
\usepackage[utf8]{inputenc}
\usepackage{hyperref} % For PDF metadata
\usepackage{enumitem} % Controlling list styling

@@

\setlist{topsep=0pt}

@@

\conferenceinfo{SIGMETRICS}{2015 Portland, Oregon USA}
\CopyrightYear{2015}
%\acmVolume{V}
%\acmNumber{N}
%\acmArticle{A}
%\acmYear{YYYY}
%\acmMonth{0}

@@

\numberofauthors{2}

% names here
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\iffalse
\alignauthor
Lars Kirkholt Melhus\\
       \affaddr{Norwegian University of Science and Technology}\\
       \affaddr{Sem Sælands vei 9}\\
       \affaddr{7491 Trondheim, Norway}\\
       \email{lars.kirkholt.melhus@gmail.com}
% 2nd. author
\alignauthor
Rune Erlend Jensen\\
       \affaddr{Norwegian University of Science and Technology}\\
       \affaddr{Sem Sælands vei 9}\\
       \affaddr{7491 Trondheim, Norway}\\
       \email{runeerle@idi.ntnu.no}
\fi
\alignauthor
Anonymous\\
       \affaddr{Anonymous University}\\
       \affaddr{Anonymous}\\
       \affaddr{Anonymous, Anonymous}\\
       \email{anonymous@anonymous.com}
% 2nd. author
\alignauthor
Anonymous\\
       \affaddr{Anonymous University }\\
       \affaddr{Anonymous}\\
       \affaddr{Anonymous, Anonymous}\\
       \email{anonymous@anonymous.com}
}

@@

% Choose from classification tree http://www.acm.org/about/class/ccs98-html
\category{C.4}{Performance of Systems}{Measurement techniques; Performance attributes}
\category{D.3.4}{Processors}{Optimization; Run-time environments}

% Terms come from a fixed list of predefined terms
\terms{Measurement, Performance}
% Free to choose keywords, but must be in alphabetical order
\keywords{4K address aliasing, BLAS, heap allocators, measurement bias, memory disambiguation, performance counters}

@@

\newpage
\subsection{Avoiding Aliasing}
Addresses of automatic variables can not be determined statically, because the location of the stack at runtime is generally unknown.
In addition to being offset by environment variables, the initial stack address can also be perturbed by for example program arguments or address randomization.
Although one can not easily know if a collision is going to happen for a given environment a priori, it is possible to change the program to account for possible alias conditions.
A proof of concept of how alias-free code could be generated in this particular case is shown in Figure \ref{lst:loopfixed}.
If the addresses do alias, a branch to an alternative but semantically equivalent code path is performed.
Calling the function recursively will effectively allocate a new set of variables a bit further down the stack, and the alias condition is avoided.

@@

\newpage
\section{Aliasing in Real Applications}
\label{sec:blas}
To investigate how real world applications can be impacted by aliasing, we look at low level numerical applications.
Basic Linear Algebra Subroutines (BLAS) is the de facto standard API for high performance linear algebra routines, with several heavily optimized implementations available~\cite{Blackford:2002:UpdatedBLAS}.
The functionality is divided into three categories:
\begin{description}
  \item[Level 1] Scalar and vector operations, such as dot product and vector addition.
  \item[Level 2] Matrix-vector operations, such as gemv for general matrix-vector multiplication.
  \item[Level 3] Matrix-matrix operations, including the widely applied gemm routine for general matrix multiplication.
\end{description}
Procedures operating on vectors, from level 1 or 2, intuitively appears the most likely to have potential for aliasing.

@@

\newpage
\section{Conclusions}
\label{sec:conclusions}
We have shown how address aliasing affects program performance under different memory contexts, and how it can explain certain cases of measurement bias.
The effect is caused by how speculative and out-of-order memory operations are handled by the CPU, which only considers the last 12 address bits to resolve conflicts between a load and previous store operations.
This phenomenon is likely present on most consumer hardware based on Intel platforms today, ranging from ``Core'' to ``Haswell'' microarchitectures.

>>>>

% paper.tex
% ACM sigmetric format
% Author: L. K. Melhus and R. E. Jensen
%\documentclass{acm_proc_article-sp}
\documentclass[10pt, conference, compsocconf]{IEEEtran}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{filecontents}
\usepackage{tikz}
\usepackage{listings}
\usepackage{mathtools}
\usepackage{underscore}
\usepackage{url}
\usepackage{lmodern} % Much clearer text
\usepackage[utf8]{inputenc}
\usepackage{hyperref} % For PDF metadata
%\usepackage{enumitem} % Controlling list styling

@@

%\setlist{topsep=0pt}

@@

%\conferenceinfo{SIGMETRICS}{2015 Portland, Oregon USA}
%\CopyrightYear{2015}
%\acmVolume{V}
%\acmNumber{N}
%\acmArticle{A}
%\acmYear{YYYY}
%\acmMonth{0}

@@

%\numberofauthors{2}

\iffalse
% names here
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author

\alignauthor
Lars Kirkholt Melhus\\
       \affaddr{Norwegian University of Science and Technology}\\
       \affaddr{Sem Sælands vei 9}\\
       \affaddr{7491 Trondheim, Norway}\\
       \email{lars.kirkholt.melhus@gmail.com}
% 2nd. author
\alignauthor
Rune Erlend Jensen\\
       \affaddr{Norwegian University of Science and Technology}\\
       \affaddr{Sem Sælands vei 9}\\
       \affaddr{7491 Trondheim, Norway}\\
       \email{runeerle@idi.ntnu.no}

\alignauthor
Anonymous\\
       \affaddr{Anonymous University}\\
       \affaddr{Anonymous}\\
       \affaddr{Anonymous, Anonymous}\\
       \email{anonymous@anonymous.com}
% 2nd. author
\alignauthor
Anonymous\\
       \affaddr{Anonymous University }\\
       \affaddr{Anonymous}\\
       \affaddr{Anonymous, Anonymous}\\
       \email{anonymous@anonymous.com}
}
\fi

\author{\IEEEauthorblockN{Authors Name/s per 1st Affiliation (Author)}
\IEEEauthorblockA{line 1 (of Affiliation): dept. name of organization\\
line 2: name of organization, acronyms acceptable\\
line 3: City, Country\\
line 4: Email: name@xyz.com}
\and
\IEEEauthorblockN{Authors Name/s per 2nd Affiliation (Author)}
\IEEEauthorblockA{line 1 (of Affiliation): dept. name of organization\\
line 2: name of organization, acronyms acceptable\\
line 3: City, Country\\
line 4: Email: name@xyz.com}
}

@@

% Choose from classification tree http://www.acm.org/about/class/ccs98-html
%\category{C.4}{Performance of Systems}{Measurement techniques; Performance attributes}
%\category{D.3.4}{Processors}{Optimization; Run-time environments}

% Terms come from a fixed list of predefined terms
%\terms{Measurement, Performance}
% Free to choose keywords, but must be in alphabetical order
\IEEEkeywords{4K address aliasing, BLAS, heap allocators, measurement bias, memory disambiguation, performance counters}

@@

%\newpage
\subsection{Avoiding Aliasing}
Addresses of automatic variables can not be determined statically, because the location of the stack at runtime is generally unknown.
In addition to being offset by environment variables, the initial stack address can also be perturbed by for example program arguments or address randomization.
Although one can not easily know if a collision is going to happen for a given environment a priori, it is possible to change the program to account for possible alias conditions.
A proof of concept of how alias-free code could be generated in this particular case is shown in Figure \ref{lst:loopfixed}.
If the addresses do alias, a branch to an alternative but semantically equivalent code path is performed.
Calling the function recursively will effectively allocate a new set of variables a bit further down the stack, and the alias condition is avoided.

@@

%\newpage
\section{Aliasing in Real Applications}
\label{sec:blas}
To investigate how real world applications can be impacted by aliasing, we look at low level numerical applications.
Basic Linear Algebra Subroutines (BLAS) is the de facto standard API for high performance linear algebra routines, with several heavily optimized implementations available~\cite{Blackford:2002:UpdatedBLAS}.
The functionality is divided into three categories:
\begin{description}
  \item[Level 1] Scalar and vector operations, such as dot product and vector addition.
  \item[Level 2] Matrix-vector operations, such as gemv for general matrix-vector multiplication.
  \item[Level 3] Matrix-matrix operations, including the widely applied gemm routine for general matrix multiplication.
\end{description}
Procedures operating on vectors, from level 1 or 2, intuitively appears the most likely to have potential for aliasing.

@@

%\newpage
\section{Conclusions}
\label{sec:conclusions}
We have shown how address aliasing affects program performance under different memory contexts, and how it can explain certain cases of measurement bias.
The effect is caused by how speculative and out-of-order memory operations are handled by the CPU, which only considers the last 12 address bits to resolve conflicts between a load and previous store operations.
This phenomenon is likely present on most consumer hardware based on Intel platforms today, ranging from ``Core'' to ``Haswell'' microarchitectures.
